---
title: "PCA Covariance - Ranking Companies"
author: "Edrei Santos"
output: html_notebook
---

  The excel file presented in the book show data relative to 12 companies and 3 variables measured in monetary currency. The variables are: Gross_Profit (ganho_bruto), Net Profit (ganho_líquido) and Total Assets (patrimônio). The goal is to rank those companies by their overall performance in those three variables.

  Before starting the analysis, lets take a look at the data:

```{r}
library(readxl)
data = read_excel("Data_Chapter_3.1_Mingoti.xlsx", sheet= 1)
print(data)
```
   We cannot do the operations in the dataset above, because there is a column of chars stating the number of the company. We need to remove this column. The information of the company number is not lost, it is stored in the row indexes.
   
```{r}
data = data[,-1] # remove the first column
print(data)
```
   Now we can finally operate. First, I want to see the descriptive stats of the data
   
```{r}
summary(data)
cat("\n")
cat("Standard Deviation of each variable: \n\n")
apply(data, 2, sd)
```
   All of those variables are deeply related, so we can use PCA to find a score that take into account the relationships between all of them. To do that, we also need the covariance matrix of the data:
   
```{r}
cov_data = cov(data)
cat("Covariance Matrix: \n\n")
print(cov_data)
```
   We will now find the eigenvectors and eigenvalues to get the orthogonality of the data.
   
```{r}
eigVal = eigen(cov_data)$values
eigVec = eigen(cov_data)$vectors

cat("Eigenvalues: \n\n")
print(eigVal)
cat("\n")
cat("Eigenvectors: \n\n")
print(eigVec)
```
  The scores of the PCA are given by the product of the data matrix and the eigenvectors. But first, I want to see how much variance is explained by each component. To find that, we need to divide the eigenvalues by the total variance, which is the sum of all eigenvalues.
```{r}
total_variance = sum(eigVal)
variance_explained = eigVal / total_variance
cat("Variance explained by each component: \n\n")
print(variance_explained)
```
  We can see that the first component explains almost 94,2% of the variance, while the other two components explain only 5,8% of the variance. I want now to check the correlation between the variables and the components to help the interpretation of what the components mean.
  
```{r}
variance_array = diag(cov_data)

# Multiplying the eigenvectors by the square root of the eigenvalues in a column-wise manner
variable_and_component_correlation = sweep(eigVec, 2, sqrt(eigVal), FUN = "*")

# Dividing by the standard deviation of the variables in a row-wise manner
variable_and_component_correlation = sweep(variable_and_component_correlation, 1, sqrt(variance_array), FUN = "/")

cat("Correlation between variables (rows) and components (columns): \n\n")
print(variable_and_component_correlation)
```
  We can now see that the first component is strongly correlated with all three variables, making it the perfect candidate for the overall performance score. 
  
  The second component is a comparison between the gross profit and total assets, while the third one is basically made of the net profit, which is not very useful because the net profit is already included in the variables. They are not very useful for what we want.
  
  Its time to get the scores.
  
```{r}

```
  
  